Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-07-06 Manuscript/manuscript.Rmd
Review

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome <!--may affect genomic regulation--> [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking; @dixon2012topological]. Likewise, evidence suggests that the boundaries that delineate TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes [@symmons2014functional]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d; @symmons2014functional] [??? This sentence and reference belong to the previous paragraph, think and move; !!!combined and moved]. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

Distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [@dixon2015chromatin; @di2017novo] [??? REFS? Carefully select from my paper notes; !!!added refs]. Recent research has identified key functional genomic annotations enriched at boundaries. Namely, the CCCTC-binding factor  (CTCF), has been shown to be associated with TAD boundaries [@ghirlando2016ctcf]. Other transcription factors including SMC1, SMC3, and RAD21 have been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Additional binding proteins, known to mediate enhancer-promoter interactions including YY1 and ZNF143, have been shown to influence the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. <!--However, there is a growing number of publically available annotated genomic features, and it is unclear if current methods are robust enough to identify all of them. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it. !!!editted]-->

Current methods used to predict TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These TAD boundary regions make up the minority class, while the non-TAD boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. A dataset is considered imbalanced if PI<0.5. <!--Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008].--> Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

2. **Overlap Percent (OP)**: For each genomic bin, the percent of overlap between the width of the region assigned to a genomic annotation and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with a particular genomic bin. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin.

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated.

@@

Class imbalance was characterized by each dataset's *proportion of imbalance* (PI), defined as the ratio of minority class observations to the sum of minority and majority class observations:

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Additional performance metrics were reported in the supplementary materials.

@@

We analyzed publically available replicate Hi-C data sets from two cell lines including GM12878 and K562 (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at four resolutions: 10 kb, 25 kb, 50 kb, and 100 kb. To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD versus non TAD boundary regions on the linear genome, labeled as a response vector **Y** (Figure 1). First we binned the linear genome according to the data resolution. We then flanked the boundaries on either side by the same number of basepairs as the previously constructed genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$).

@@

We first sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were found to be specific to one of the two cell lines (Figure 2B). It was found that 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, across each resolution respectively. Similarly, for K562, it was found that 27%, 27%, 29%, and 33% of TAD boundary regions were cell line specific for each resolution specifically. This indicates a necessity for computationally specific models in predicting TAD boundary formations.

@@

We then investigated the proportional class imbalance (PI) featured at each resolution and among each cell line. The PI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to classification.

@@

A total of 107 and 152 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line respectively. To establish the feature space used for classification, we constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 3A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 3B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is currently unexplored (Figure 3C). Log2-transformation of distance was used as a normalization technique in order to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classifier performance in class imbalance settings.

@@

In order to identify the optimal combination of data resolution, predictor type, and re-sampling technique necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 4). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line ($L = \{GM12878, K562\}$). Within each cell line we evaluated 4 data resolutions ($R = \{10 kb, 25 kb, 50 kb, 100 kb\}$). We then considered 3 separate predictor types as the feature space ($P = \{OC, OP, Distance\}$). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling ($B = \{None, RUS, ROS, SMOTE\}$). Each combination of the above cell line specific parameters was implemented on chromosome specific sections of the linear genome (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. For each combination of inputs, we performed 10-fold cross-validation to tune for the number of features to consider at each node of the random forest classifier. Each optimally tuned model was then validated on the same separate testing set for unbiased comparisons. Performances were evaluated using MCCs and aggregated across each chromosome by averaging. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

@@

We first evaluated the least granular cell line specific random forest classifiers, those were ones implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 5 A-B (red)). We then assessed models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no clear improvement (Figure 5 C-D (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance, regardless of cell line and resolution. However, when cell line specific classifiers were implemented using distance-type predictors and no re-sampling, we found that there was a significant improvement in MCC compared to the other two predictor types (Figure 5 E-F (red)).

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models, within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that cell line specific models built simply on TFBS out performed other cell line and annotation-specific models at all resolutions (Figure 7). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations for both cell lines. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal.

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

@@

[Put this section last]

@@

We then investigated the spatial patterns of the most predictive TFBS around TAD boudaries by characterizing the distance profiles between the classes of genomic bins. For GM12878, at 10 kb resolution, we found that SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were all statistically significantly closer to bins that overlapped with flanked TAD boudaries than those that did not (Figure 8). Although still statistically significantly different, the distance profiles between genomic class were found to be much more similar at lower resolutions (Supplementary Figure 9). For the K562 cell line, we reported the distance profiles of the same transcription factor binding sites, excluding the YY1sc281, due to its unavailability in the UCSC Genome Browser. These cell line specific distance profiles were again found to be significantly different between genomic class, at all resolutions (Supplementary Figure 10).

@@

In this study we developed an ensemble framework to systematically compare model performances for different combinations of re-sampling and feature engineering techniques aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations. We provide direct comparisons of cell line specific random forest models built on a variety of different data dependent characteristics, including 4 contact matrix resolutions ($\{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$), 3 predictor-types ($\{\textit{OC}, \textit{OP}, \textit{Distance}\}$), and 4 re-sampling techniques ($\{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$), for both the GM12878 and K562 cell lines.

The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Using the framework we demonstrated that random forest models built on imbalanced data exhibited poor performance, regardless of cell line, resolution, or predictor type. In our framework, we opted to leave the test set as a completely untouched holdout set, only performing the re-sampling, and subsequent learning, on the training set. This was done so as to incur the least amount of bias and so that all conclusions could be made soley between model characteristics. As a result, we showed that Matthews Correlation Coefficient was much more sensitive to the imbalance present in the test set, compared to more popular metrics like AUROC, which attain artificially high values by correctly predicting majority class samples [?? I may be mistaken but I don't see where this is proven ??]. When comparing re-sampling techniques, we found that SMOTE re-sampling outperformed models built on both simple ROS or RUS, implying that, in fact, a combination of both random under-sampling and random over-sampling is preferred in TAD boundary prediction. Furthermore, the most improvement [?? Over what ??] was seen when combining SMOTE resampling on distance-type features when building random forest classifiers. These two characteristics exhibited consistently better performance results across both the GM12878 and K562 cell lines. This confirmed our hypothesis that the most granular and spatially informative representation of the relationship between TAD boundaries and annotations resulted in the best performing model.

@@

We also observed that, while the best set of data characteristics were consistent across resolutions, performances of classifiers built on said characteristics tended to decline as resolution decreased. For both cell lines, top performance was achieved at the 10 kb data resolution, while the worst performance was obtained at the 50 kb and 100 kb resolution for the GM12878 and K562 cell line respectively. This suggests that TAD boundary prediction is intimately related to the resolution that boundaries are called on. This result can be attributed to one of two factors in the scheme of our framework. First, at higher resolutions, by model construction genomic bins were smaller, thereby resulting in more accurate profiles for the genomic annotations [?? Don't understand this sentence ??]. Secondly, because we are mainly concerned with TAD boundaries, and not the hierarchical structure of domains as a whole, the sparsity of contact matrices at higher resolutions does not impair boundary prediction [?? Doesn't this contradict your results about low resolution causing worse boundary prediction ??]. Conversely, lower resolutions imply larger bin windows and lead to less specificity in boundary location on the linear genome, resulting in worse performances. Given the trend in performances across resolution, future research is necessary in determining the optimal resolution to use when predicting TAD boundaries [?? Might want to qualify the statement in results about 10kb data being best by saying that it only applies to your data ??].

Given the formulation of our framework, we were not able to make direct comparisons between cell lines. This was due to the fact that we opted to use cell line specific annotations, and there existed some annotations that were available on the K562 cell line that were not available to GM12878, and vice versa. However, we were able to confirm that final models built on the optimal set of data characteristics identified consistent and well known proteins related to the 3D structure of chromatin, for both cell lines. Our results indicated that the zinc finger protein CTCF, as well components of the cohesin complex, SMC3 and RAD21, were found to be the key molecular drivers most predictive of TAD boundary formation, consistent across cell line and resolution. In addition, models derived from our framework recognized known mediaters of long range enhancer-promoter interactions, the ZNF143 and YY1 transcription factors, as being highly predictive of TAD boundaries.

>>>>

The advent of various genome-wide sequencing technologies, such as high-throughput conformation capture (notably Hi-C), have revealed the spatial organization of the human genome <!--may affect genomic regulation--> [@lieberman2009comprehensive]. Analyses have shown that the genome is folded into distinct compartments termed topologically associating domains (TADs) [@dixon2012topological]. Given their high degree of conservation between cell types and species, it has been proposed that TADs represent the fundamental building blocks of organization, giving the genome its 3D structure [@rocha2015breaking; @dixon2012topological]. Likewise, evidence suggests that the boundaries that delineate TADs may play a role in restricting the function of certain genomic elements such as enhancers, thereby impacting the transcription of genes [@symmons2014functional]. Even more so, loci within a domain have been shown to interact more frequently with one another, exhibiting similar patterns of chromatin modifications, and long-range contact patterns [@rao20143d; @symmons2014functional]. Furthermore, studies have shown that changing the 3D structure of the DNA, causing disruptions in these domains, can lead to adverse outcomes and diseases like cancer [@hnisz2016activation; @Flavahan2016; @song2017ctcf]. Therefore, it has become increasingly important to be able to identify the key molecular drivers of the formation of TAD boundaries in order to further our understanding of the human genome.

Distinct patterns in epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify which epigenomic features are most predictive of TAD boundaries [@dixon2015chromatin; @di2017novo]. Recent research has identified key functional genomic annotations enriched at boundaries. Namely, the CCCTC-binding factor  (CTCF), has been shown to be associated with TAD boundaries [@ghirlando2016ctcf]. Other transcription factors including SMC1, SMC3, and RAD21 have been shown to colocalize near TAD boundaries [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Together, these components are hypothesized to be what causes loops in the DNA, contributing to higher order chromatin organization. Additional binding proteins, known to mediate enhancer-promoter interactions including YY1 and ZNF143, have been shown to influence the delineation of topological domains [@weintraub2017yy1; @gonzalez2017znf143; @heidari2014genome]. <!--However, there is a growing number of publically available annotated genomic features, and it is unclear if current methods are robust enough to identify all of them. [??? This is very weak and undefined. You started with mentioning prediction - finish with unsolved problems with it. !!!editted]--> [??? What's the summary/conclusion/problem?]

Current methods used to predict [??? you never introduced prediction problem] TAD boundaries rely on the use of classification models to draw associations between functional annotations and genomic regions that include TAD boundaries. However, many key characteristics of 3D genomic data remain unexplored, potentially limiting performance of predictive modeling. Specifically, the minority of TAD boundaries as compared with the rest of the genome creates proportionally small TAD boundary regions versus non-TAD boundary regions. These TAD boundary regions make up the minority class, while the non-TAD boundary regions make up the remaining majority class. The large disparity is sizes between minority and majority classes creates imbalanced data, as can be characterized by the *proportion of imbalance* (PI). Here, PI is defined as the ratio of the set minority class observations to the sum of both sets of minority and majority class observations. A dataset is considered imbalanced if PI<0.5. [??? Not really. Poor metric choice.] <!--Specifically, the minority of TAD boundaries as compared with the rest of the genome creates heavily imbalanced classes, characterized by ratios of positive/negative classes used to train the models. Here, the imbalance ratio is defined as the ratio of majority classes (regions not including TAD boundaries) to minority classes (regions including TAD boundaries) [@Orriols-Puig2008].--> Classifiers implemented on heavily imbalanced datasets are known to exhibit poor performances as the learned model is biased towards the majority class [@lusa2010class; @chen2013influence; @jeni2013facing]. Furthermore, the use of different levels of association between TAD boundaries and genomic features (number, or proportion of overlaps, or distance) as predictors remain underexplored. Lastly, it remains unclear how increased resolution of Hi-C data might affect the performance of predictive models.

@@

2. **Overlap Percent (OP)**: For each genomic bin, the percent of overlap between the width of the region assigned to a genomic annotation and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with a particular genomic bin. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. [??? Unclear.]

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter as $\textit{distance}$) [??? Learn Markdown!], from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated.

@@

Class imbalance was characterized by each dataset's *proportion of imbalance* (PI), [??? Why? It is bounded [0, 0.5], never seen it is used.] defined as the ratio of minority class observations to the sum of minority and majority class observations:

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Additional performance metrics were [??? are?] reported in the supplementary materials. [??? Where is it?]

@@

We analyzed publically available replicate Hi-C data sets from two cell lines including GM12878 and K562 (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at four resolutions: 10 kb, 25 kb, 50 kb, and 100 kb. To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD versus non TAD boundary regions on the linear genome, labeled as a response vector **Y** (Figure 1). First, we binned the linear genome according to the data resolution. We then flanked the boundaries on either side by the same number of basepairs as the previously constructed genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$).

@@

We sought to compare the cell-line specificity of TAD boundary regions. Between the two cell lines, a significant number of TAD boundary regions were found to be specific to one of the two cell lines (Figure 2B). It was found that 45%, 38%, 37%, and 36% of TAD boundary regions were specific to the GM12878 cell line, across each resolution [??? spell out resolutions exactly.] respectively. Similarly, for K562, it was found that 27%, 27%, 29%, and 33% of TAD boundary regions were cell line specific for each resolution specifically. This indicates a necessity for computationally specific models [??? Do not understand] in predicting TAD boundary formations.

@@

We then investigated the proportional class imbalance (PI) featured at each resolution and among each cell line. The PI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions. The proportional imbalance was found to be invariant to resolution, remaining consistently low for both cell lines, with very little variability between chromosomes (Figure 3C). Such low proportions of imbalance gave precedence for the implementation of re-sampling prior to classification.  [??? It is low because the upper bound is 0.5. It is an unconventional measure.]

@@

A total of 107 and 152 cell type specific functional genomic annotations derived from ChIP-Seq experiments (See Methods) for the GM12878 and K562 cell line, respectively. [??? Is that so? How many are common?] To establish the feature space used for classification, we constructed three different predictor types for each functional genomic annotation. Counts of overlaps (**overlap count**, OC) is the most commonly used type of association between genomic elements (Figure 3A). Percents of overlaps (**overlap percent**, OP) represents a more precise measure of association as it considers the proportion of a bin overlapped by a genomic annotation (Figure 3B). **Distance** between the centers of bins and genomic annotations quantifying the spatial relationship between genomic elements is currently unexplored (Figure 3C). Log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figures 1-2). These types of genomic associations provide a granular way to investigate classifier performance in class imbalance settings.

@@

In order to identify the optimal combination of data resolution, predictor type, and re-sampling technique necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 4). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line ($L = \{GM12878, K562\}$). Within each cell line we evaluated 4 data resolutions ($R = \{10 kb, 25 kb, 50 kb, 100 kb\}$). We then considered 3 separate predictor types as the feature space ($P = \{OC, OP, Distance\}$). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling ($B = \{None, RUS, ROS, SMOTE\}$). Each combination of the above cell line specific parameters was implemented on chromosome specific sections of the linear genome (omitting chromosome 9). Letting $|X|$ denote the cardinality [??? TBD] of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. For each combination of inputs, we performed 10-fold cross-validation to tune for the number of features to consider at each node of the random forest classifier. Each optimally tuned model was then validated on the same separate testing set for unbiased comparisons. Performances were evaluated using MCCs and aggregated across each chromosome by averaging. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

@@

We first evaluated the least granular cell line specific random forest classifiers [??? Unclear], those were ones implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 5 A-B (red)). We then assessed models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no clear improvement (Figure 5 C-D (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance, regardless of cell line and resolution. However, when cell line specific classifiers were implemented using distance-type predictors and no re-sampling, we found that there was a significant improvement in MCC compared to the other two predictor types (Figure 5 E-F (red)).

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models [??? Never defined, what is this?], within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that cell line specific models built simply on TFBS outperformed other cell line and annotation-specific models at all resolutions (Figure 7). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations for both cell lines. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal.

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. [??? Where do you show it?] In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

@@

@@

We then investigated the spatial patterns of the most predictive TFBS around TAD boudaries by characterizing the distance profiles between the classes of genomic bins. For GM12878, at 10 kb resolution, we found that SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were all statistically significantly closer to bins that overlapped with flanked TAD boudaries than those that did not (Figure 8). [??? not there] Although still statistically significantly different, the distance profiles between genomic class were found to be much more similar at lower resolutions (Supplementary Figure 9). For the K562 cell line, we reported the distance profiles of the same transcription factor binding sites, excluding the YY1sc281, due to its unavailability in the UCSC Genome Browser. These cell line specific distance profiles were again found to be significantly different between genomic class, at all resolutions (Supplementary Figure 10).

@@

In this study we developed an ensemble framework to systematically compare model performances for different combinations of re-sampling and feature engineering techniques aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations. We provide direct comparisons of cell line specific random forest models built on a variety of different data dependent characteristics, including 4 contact matrix resolutions ($\{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$), 3 predictor-types ($\{\textit{OC}, \textit{OP}, \textit{Distance}\}$), and 4 re-sampling techniques ($\{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$), for both the GM12878 and K562 cell lines. [??? Learn Markdown] [??? What's the summary of your work?]

The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Using the framework we demonstrated that random forest models built on imbalanced data exhibited poor performance, regardless of cell line, resolution, or predictor type. In our framework, [??? Too much "framework" mentioning] we opted to leave the test set as a completely untouched holdout set, only performing the re-sampling, and subsequent learning, on the training set. This was done so as to incur the least amount of bias and so that all conclusions could be made soley between model characteristics. As a result, we showed that Matthews Correlation Coefficient was much more sensitive to the imbalance present in the test set, compared to more popular metrics like AUROC, which attain artificially high values by correctly predicting majority class samples [??? I don't see results for it]. When comparing re-sampling techniques, we found that SMOTE re-sampling outperformed models built on both simple ROS or RUS, implying that, in fact, a combination of both random under-sampling and random over-sampling is preferred in TAD boundary prediction. Furthermore, the most improvement [??? In what?] was seen when combining SMOTE resampling on distance-type features when building random forest classifiers. These two characteristics exhibited consistently better performance results across both the GM12878 and K562 cell lines. This confirmed our hypothesis that the most granular and spatially informative representation of the relationship between TAD boundaries and annotations resulted in the best performing model.

@@

We also observed that, while the best set of data characteristics were consistent across resolutions, performances of classifiers built on said characteristics tended to decline as resolution decreased. For both cell lines, top performance was achieved at the 10 kb data resolution, while the worst performance was obtained at the 50 kb and 100 kb resolution for the GM12878 and K562 cell line respectively. This suggests that TAD boundary prediction is intimately related to the resolution that boundaries are called on. This result can be attributed to one of two factors in the scheme of our framework. First, at higher resolutions, by model construction genomic bins were smaller, thereby resulting in more accurate profiles for the genomic annotations [?? Don't understand this sentence ??]. Secondly, because we are mainly concerned with TAD boundaries, and not the hierarchical structure of domains as a whole, the sparsity of contact matrices at higher resolutions does not impair boundary prediction [?? Doesn't this contradict your results about low resolution causing worse boundary prediction ??]. Conversely, lower resolutions imply larger bin windows and lead to less specificity in boundary location on the linear genome, resulting in worse performances. Given the trend in performances across resolution, future research is necessary in determining the optimal resolution to use when predicting TAD boundaries [?? Might want to qualify the statement in results about 10kb data being best by saying that it only applies to your data ??]. [??? This paragraph is a mess]

Given the formulation of our framework, we were not able to make direct comparisons between cell lines. This was due to the fact that we opted to use cell line specific annotations, and there existed some annotations that were available on the K562 cell line that were not available to GM12878, and vice versa. [??? Did they provide additional advantage?] However, we were able to confirm that final models built on the optimal set of data characteristics identified consistent and well known proteins related to the 3D structure of chromatin, for both cell lines. Our results indicated that the zinc finger protein CTCF, as well components of the cohesin complex, SMC3 and RAD21, were found to be the key molecular drivers most predictive of TAD boundary formation, consistent across cell line and resolution. In addition, models derived from our framework recognized known mediaters of long range enhancer-promoter interactions, the ZNF143 and YY1 transcription factors, as being highly predictive of TAD boundaries.
